---
title: "Simple NNGP"
author: "Elizabeth Chun"
date: "2024-11-11"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
# library(FRK)

# load packages
library(tidyverse)
library(tictoc)

library(geoR)
library(BRISC)

# plotting
library(sf)
library("rnaturalearth")
```


# load data

```{r}
years <- c("2019", "2020", "2021", "2022", "2023")
data_dir <- "final_project/AirQualityData/"

years_list <- list()

for (yr in years) {
  yr_names <- list.files(paste0(data_dir, yr), full.names = T)
  yr_list <- list()
  
  for (file in yr_names) {
    file_df <- read.csv(file, header = T)
    
    # Generate a unique name for each layer
    name <- gsub(".csv", "", file)  # "Replace" the file extension with ""
    name <- gsub(paste0(data_dir, yr, "/"), "", name)
    
    yr_list[[name]] <- file_df
  }
  
  years_list[[yr]] <- yr_list
}
```

```{r}
gases = names(years_list$'2023')
twothree_list = list()

for (gas in gases) {
  current = years_list$'2023'[[gas]]
  ncols = ncol(current)
  current = current[, c(1:6, (ncols-1):ncols)]
  colnames(current)[5] = 'Daily_Max'
  
  current['pollutant'] = gas
  
  twothree_list[[gas]] = current
  
}

twothree = do.call('rbind', twothree_list)
```

```{r}
aug02 = twothree %>%
  filter(Date == '08/02/2023') %>%
  filter(pollutant == 'NO2')
rownames(aug02) = 1:nrow(aug02)
```

```{r}
aug02_sf = st_as_sf(aug02, coords = c("Site.Longitude", "Site.Latitude"), crs = 4326)
bbox_no2 = st_bbox(aug02_sf)

aug02_sf %>%
  ggplot() + 
    base_map(bbox_no2, increase_zoom = 1, basemap = 'mapnik') +
    geom_sf(aes(color = Daily_Max), alpha = 0.75, size = 5) +
  scale_color_gradient2(low = 'darkgreen', mid = 'orange', high = "red", 
                       limits = c(0, 60), na.value = NA, midpoint = 30)
```


# NNGP

```{r data splitting}
### data splitting
set.seed(27)

# get 80/20 train test split
train_idx = sample(nrow(aug02), 0.8*nrow(aug02))
train_all = aug02[train_idx, ]
test_all = aug02[-train_idx, ]
```

```{r}

st_as_sf(train_all, coords = c("Site.Longitude", "Site.Latitude"), crs = 4326) %>%
  ggplot() + 
    base_map(bbox_no2, increase_zoom = 1, basemap = 'mapnik') +
    geom_sf(aes(color = Daily_Max), alpha = 0.75, size = 5) +
    scale_color_gradient2(low = 'darkgreen', mid = 'orange', high = "red", 
                         limits = c(0, 60), na.value = NA, midpoint = 30)
```

```{r}

st_as_sf(test_all, coords = c("Site.Longitude", "Site.Latitude"), crs = 4326) %>%
  ggplot() + 
    base_map(bbox_no2, increase_zoom = 1, basemap = 'mapnik') +
    geom_sf(aes(color = Daily_Max), alpha = 0.75, size = 5) +
    scale_color_gradient2(low = 'darkgreen', mid = 'orange', high = "red", 
                         limits = c(0, 26), na.value = NA, midpoint = 20)
```


## 3.2 Nearest neighbors GP

Now I will fit an NNGP on the entire training set with 10 neighbors using the BRISC package.

```{r fit nngp}
fit_nngp = BRISC_estimation(coords = as.matrix(train_all[, 7:8]), 
                            x = cbind(1, as.matrix(train_all[, 7:8])), y = train_all$Daily_Max, 
                            cov.model = 'exponential', n.neighbors = 10)
print(fit_nngp$Beta)
print(fit_nngp$Theta)
```



# 4. Model results and predictions

## 4.2 Nearest neighbors GP

For the NNGP, here are the estimated parameters. Note that BRISC reports $\phi$ as the reciprocal of the standard definition so I have taken $\frac{1}{\phi}$.

$\hat{\beta}$'s = [`r fit_nngp$Beta`]  
$\hat{\sigma}^2$ = `r fit_nngp$Theta[1]`  
$\phi$ = `r 1/fit_nngp$Theta[3]`  
$\hat{\tau}^2$ = `r fit_nngp$Theta[2]`    

 9.311698 with 10 nn

```{r nngp prediction}
preds_nngp = BRISC_prediction(fit_nngp, coords.0 = as.matrix(test_all[, 7:8]), 
                              X.0 = cbind(1, as.matrix(test_all[, 7:8])))
rmse_nngp = sqrt(mean((test_all$Daily_Max - preds_nngp$prediction)^2))
mape = mean(abs(test_all$Daily_Max - preds_nngp$prediction)/test_all$Daily_Max*100)

resids = test_all$Daily_Max - preds_nngp$prediction

print(rmse_nngp)
print(mape)

hist(resids)
hist(preds_nngp$prediction)
```

```{r}
# plot random predictions map

pred_df_rand = test_all[, c('Site.Longitude', 'Site.Latitude')] %>%
  mutate(
    no2_ppm_pred = preds_nngp$prediction
  ) 
pred_df_rand_sf = st_as_sf(pred_df_rand, coords = c('Site.Longitude', 'Site.Latitude'), crs = 4326)

ggplot(pred_df_rand_sf) + 
  base_map(bbox_no2, increase_zoom = 1, basemap = 'mapnik') +
  geom_sf(aes(color = no2_ppm_pred), alpha = 0.75, size = 5) +
  scale_color_gradient2(low = 'darkgreen', mid = 'orange', high = "red", 
                       limits = c(0, 26), na.value = NA, midpoint = 20)
```

# predict on rough grid

```{r}
si <- expand.grid(lon =seq(from = -106, to = -93, by = 0.5), 
                  lat = seq(from = 25, to = 36, by = 0.5))
si_sf = st_as_sf(si, coords = c('lon', 'lat'), crs = 4326)

tx_xform = sf::st_transform(tx, crs = 4326)
texas_si = si_sf[tx_xform, ]

texas_si_final = texas_si %>% 
  st_drop_geometry() %>%  
  bind_cols(st_coordinates(texas_si)) %>%
  select(lat = Y, lon = X)
```


```{r nngp prediction}
preds_grid = BRISC_prediction(fit_nngp, coords.0 = as.matrix(texas_si_final), 
                              X.0 = cbind(1, as.matrix(texas_si_final)))

hist(preds_grid$prediction)
```

```{r}
# plot random predictions map

pred_df_rand = texas_si_final %>%
  mutate(
    no2_ppm_pred = preds_grid$prediction
  ) 
pred_df_rand_sf = st_as_sf(pred_df_rand, coords = c('lon', 'lat'), crs = 4326)

ggplot(pred_df_rand_sf) + 
  base_map(bbox_no2, increase_zoom = 1, basemap = 'mapnik') +
  geom_sf(aes(color = no2_ppm_pred), alpha = 1, size = 3) +
  scale_color_gradient2(low = 'darkgreen', mid = 'orange', high = "red", 
                       limits = c(0, 26), na.value = NA, midpoint = 20)
```

