---
title: "STAT647 Project Data Wrangling"
author: "Sophia Lazcano" 
output: pdf_document
---

# **Libraries and Reading In Data**

```{r libraries}
library(readxl)
library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(ggplot2)
```

```{r dataSetup, cache=TRUE}
year <- "2019"
no2_df <- read.csv("./data/annual_conc_by_monitor_2019.csv", header = T,
                   colClasses = c(State.Code = "character",
                                  County.Code = "character"))
no2_data <- no2_df %>%
  filter((Parameter.Name == "Nitrogen dioxide (NO2)") &
           !(State.Code %in% c("02","15","72")) &
           (Pollutant.Standard == "NO2 1-hour 2010") &
           (Metric.Used == "Daily Maximum 1-hour average")) %>%
   select(State.Code, County.Code, Latitude, Longitude, Datum, Sample.Duration,
          Metric.Used, Units.of.Measure, Observation.Count, Observation.Percent,
          Arithmetic.Mean, Arithmetic.Standard.Dev, X50th.Percentile) %>%
  mutate(GEOID = paste0(State.Code, County.Code))
```


# **Data Processing**

### Feature Engineering and NA Values


```{r checkingNA}
# Check for NA values
print("Number of NA values by year and pollutant")
full_df_missing <- no2_data[apply(no2_data,
                                  1,
                                  function(x) any(is.na(x))
                                  ),
                            ]
# Showing rows with missing data
print(paste("NA values:", nrow(full_df_missing),
            "Rows:", dim(no2_data)[1],
            "Columns:", dim(no2_data)[2]
            ))
```

No NA values!

```{r moreFeatures, cache=TRUE}
# Add more feature engineering code here
```

### EDA 
```{r}
# Add EDA code here
```


### Adding Census Features

```{r censusSetup}
# Setup our U.S. Census Bureau Data API Key
options(tigris_use_cache = TRUE)
census_api_key("####") # Add API key here
```

```{r gettingCensusData, cache=TRUE}
yr_int <- as.integer(year) # 2019 queries

# Total Population
population <- get_acs("county", year = yr_int,
                   variables = "B01001_001E") %>%
  select(GEOID, estimate) %>% rename(total_population = estimate)
# Transportation
## Public
transit <- get_acs("county", year = yr_int,
                   variables = "B08006_008") %>%
  select(GEOID, estimate) %>% rename(total_public_transit = estimate)
## Car/truck/van alone
transit <- get_acs("county", year = yr_int,
                   variables = "B08006_003") %>%
  select(GEOID, estimate) %>% rename(total_private_vehicle = estimate) %>%
  right_join(transit, by = "GEOID")
## Car/truck/van carpool
transit <- get_acs("county", year = yr_int,
                   variables = "B08006_004") %>%
  select(GEOID, estimate) %>% rename(total_carpool_vehicle = estimate) %>%
  right_join(transit, by = "GEOID")
## No vehicle
transit <- get_acs("county", year = yr_int,
                   variables = "B08014_002") %>%
  select(GEOID, estimate) %>% rename(total_no_vehicle = estimate) %>%
  right_join(transit, by = "GEOID")

census_data <- population %>%
  left_join(transit, by = "GEOID")
```

```{r addingCensusData, cache=TRUE}
# Adding the Census data to the main df
no2_data_no_temp <- no2_data %>% left_join(census_data, by = "GEOID")
```


### Adding Temp Data

```{r gettingClimateData, cache=TRUE}
# cooling <- read.table("./data/cooling_days_data.txt", header = F)
# heating <- read.table("./data/heating_days_data.txt", header = F)
# temp_avg <- read.table("./data/temp_avg_data.txt", header = F)
# temp_max <- read.table("./data/temp_max_data.txt", header = F)
# temp_min <- read.table("./data/temp_min_data.txt", header = F)
```

```{r settingClimateNames}
# colnames(cooling) <- c("Code", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12") 
# colnames(heating) <- c("Code", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12") 
# colnames(temp_avg) <- c("Code", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12") 
# colnames(temp_max) <- c("Code", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12") 
# colnames(temp_min) <- c("Code", "01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12") 
```

```{r getTXClimateData}
# # Filter for TX data
# cooling_df <- cooling %>% mutate(Code = as.character(Code)) %>% filter(str_detect(Code, "^48"))
# heating_df <- heating %>% mutate(Code = as.character(Code)) %>% filter(str_detect(Code, "^48"))
# temp_avg_df <- temp_avg %>% mutate(Code = as.character(Code)) %>% filter(str_detect(Code, "^48"))
# temp_max_df <- temp_max %>% mutate(Code = as.character(Code)) %>% filter(str_detect(Code, "^48"))
# temp_min_df <- temp_min %>% mutate(Code = as.character(Code)) %>% filter(str_detect(Code, "^48"))
# 
# # Filter from 2019 - 2023 and make County GEOID and Year columns
# cooling_df <- cooling_df %>%
#   filter(str_detect(Code, "2019$")
#          | str_detect(Code, "2020$") | str_detect(Code, "2021$")
#          |str_detect(Code, "2022$") | str_detect(Code, "2023$")) %>%
#   mutate(GEOID = substr(Code, 1, 5), Year = as.factor(substr(Code, 8, 11))) %>%
#   select(-Code)
# 
# heating_df <- heating_df %>% 
#   filter(str_detect(Code, "2019$")
#          | str_detect(Code, "2020$") | str_detect(Code, "2021$")
#          |str_detect(Code, "2022$") | str_detect(Code, "2023$")) %>%
#   mutate(GEOID = substr(Code, 1, 5), Year = as.factor(substr(Code, 8, 11))) %>%
#   select(-Code)
# 
# temp_avg_df <- temp_avg_df %>% 
#   filter(str_detect(Code, "2019$")
#          | str_detect(Code, "2020$") | str_detect(Code, "2021$")
#          |str_detect(Code, "2022$") | str_detect(Code, "2023$")) %>%
#   mutate(GEOID = substr(Code, 1, 5), Year = as.factor(substr(Code, 8, 11))) %>%
#   select(-Code)
# 
# temp_max_df <- temp_max_df %>% 
#   filter(str_detect(Code, "2019$")
#          | str_detect(Code, "2020$") | str_detect(Code, "2021$")
#          |str_detect(Code, "2022$") | str_detect(Code, "2023$")) %>%
#   mutate(GEOID = substr(Code, 1, 5), Year = as.factor(substr(Code, 8, 11))) %>%
#   select(-Code)
# 
# temp_min_df <- temp_min_df %>% 
#   filter(str_detect(Code, "2019$")
#          | str_detect(Code, "2020$") | str_detect(Code, "2021$")
#          |str_detect(Code, "2022$") | str_detect(Code, "2023$")) %>%
#   mutate(GEOID = substr(Code, 1, 5), Year = as.factor(substr(Code, 8, 11))) %>%
#   select(-Code)
```






### Saving Data

```{r}
save(agg_years, pollutants, years, tx, agg_years_sf, pol_units, yearly_cenus_data,
     pol_val_col_names, years_list_prep, years_19_22, file = "processed_data.RData")
```


I want to aggregate my data by year averages of pollutant values (in column 1) and air quality index values (called Daily.AQI.Value), but using this code, I go from 48 unique longitude and latitude pairs to 12. How can I modify this code to do what I want in terms of columns returned but not lose my information of 




# **Combining Census, Climate, and Pollution Data**

```{r}
load("processed_data.RData")
load("climate_data_monthly_county_level.RData")
```

```{r}
complete_data <- list() # From years 2019-2022
for (yr in years_climate) {
  for (pol in pollutants) {
    pollution_data <- agg_years[[yr]][[pol]]
    
    census_data <- yearly_cenus_data[[yr]] %>%
      mutate(perc_no_vehicle = total_no_vehicle / total_population,
             perc_carpool_vehicle = total_carpool_vehicle / total_population,
             perc_private_vehicle = total_private_vehicle / total_population,
             perc_public_transit = total_public_transit / total_population) %>%
      select(-c(total_no_vehicle, total_carpool_vehicle,
                total_private_vehicle, total_public_transit))
    
    weather_data <- climate_data[[yr]] %>%
      select(-c(Longitude, Latitude)) %>% st_drop_geometry()
    
    full_yr_data <- pollution_data %>% left_join(census_data, by = "GEOID") %>%
      left_join(weather_data, by = c("GEOID", "Month"))
    
    complete_data[[yr]][[pol]] <- full_yr_data
  }
}
```

```{r}
# Save complete data
save(complete_data, years_climate, pollutants, climate_vars, month_chars,
     agg_years, yearly_cenus_data, climate_data, years_list_prep,
     tx, pol_val_col_names, pol_units, county_polys, county_points, county_points_coords,
     file = "full_data.RData")
```

## Test data

```{r}
load("full_data.RData") # old
```


