---
title: "STAT647 Project Data Modeling GAMM4"
author: "Sophia Lazcano" 
output: pdf_document
---

# **Libraries and Reading In Data**

```{r setup}
library(tidyverse)
library(sf)
library(mgcv)
library(HRW)
library(gamm4)
library(brms)
library(GGally)
set.seed(1)
```

```{r dataLoadIn, cache=TRUE}
load("NO2_CONUS_full_data.RData")
```

```{r dataSetup, cache=TRUE}
# Choose covariates
all_data <- no2_data_sf %>%
  mutate(perc_no_vehicle = total_no_vehicle / total_population,
         perc_carpool_vehicle = total_carpool_vehicle / total_population,
         perc_private_vehicle = total_private_vehicle / total_population,
         perc_public_transit = total_public_transit / total_population
         ) %>%
  select(State.Code, GEOID, Latitude, Longitude, total_population, ppt, tmax,
         tmean, tmin, perc_no_vehicle, perc_carpool_vehicle,
         perc_private_vehicle, perc_public_transit, X50th.Percentile)

# Data splitting
### 80-20% train-test split
train_idx <- sample(nrow(all_data), 0.8*nrow(all_data))
train_df <- all_data[train_idx, ] %>% st_drop_geometry()
test_df <- all_data[-train_idx, ] %>% st_drop_geometry()
```


# **Modeling**

Our data is stored as test-train split data frames

## **Bayesian Hierarchical Model**

### NO2 to Test Performance

```{r fitBHM, cache=TRUE}
# Fit Bayesian Hierarchical Model
bhm_no2_model <- brm(
  formula = X50th.Percentile ~ 
    (1 | GEOID) +  # Random intercept for counties
    s(Latitude, Longitude) +  # Spatial smooth (spline-based, not a spline itself)
    as.factor(State.Code) +
    perc_carpool_vehicle + perc_private_vehicle + total_population +
    s(ppt) + s(tmax) + s(tmin) + s(tmean),  # Covariates
  family = gaussian(),  # Gaussian for continuous response
  data = train_df,
  chains = 4, # Num of Markov Chains
  iter = 6000, # For each chain, higher means chains can mix, r-hat at 2000: 4.55
  cores = 5, # My laptop can handle this lol
  control = list(adapt_delta = 0.98, max_treedepth = 12)  # Improve convergence
) # tree depth default at 10
```

```{r}
# Save model in case of time-out
save(bhm_no2_model, file = "bhm_no2_chain4_iter6000_depth12_delta98.RData")
```

```{r}
# Check model summary
summary(bhm_no2_model)

# Check convergence diagnostics
plot(bhm_no2_model)
```

```{r}
# Prediction on Test Data
# Generate predictions for test data
test_df$predicted <- predict(bhm_no2_model,
                             newdata = (test_df %>%
                                          mutate(State.Code =
                                                   as.factor(State.Code)) %>%
                                          filter(State.Code != "28")
                                          ))[, "Estimate"]

# Evaluate Predictions
rmse <- sqrt(mean((test_df$X50th.Percentile - test_df$predicted)^2, na.rm = TRUE))
mae <- mean(abs(test_df$X50th.Percentile - test_df$predicted), na.rm = TRUE)

cat("RMSE:", rmse, "\nMAE:", mae)
```



## **GAM**

```{r initGAM, cache=TRUE}
gam_pol_spline_coords <- gamm4(X50th.Percentile ~
                                 s(Latitude) + s(Longitude) +
                                 as.factor(State.Code) +
                                 perc_carpool_vehicle + perc_private_vehicle +
                                 total_population +
                                 s(ppt) + s(tmax) + s(tmin) + s(tmean),
                               random = ~(1|GEOID),
                               data = train_df)
gam_pol_linear_coords <- gamm4(X50th.Percentile ~
                                 Latitude + Longitude +
                                 as.factor(State.Code) +
                                 perc_carpool_vehicle + perc_private_vehicle +
                                 total_population +
                                 s(ppt) + s(tmax) + s(tmin) + s(tmean),
                               random = ~(1|GEOID),
                               data = train_df)
```

```{r initCompare, cache=TRUE}
anova(gam_pol_spline_coords$mer, gam_pol_linear_coords$mer)
```

Since our p-value is 1, then the ANOVA test favors the simpler model with 61 parameters instead of the spline model with 63 parameters.

```{r nextCompare, cache=TRUE}
gam_pol_linear_ppt <- gamm4(X50th.Percentile ~
                                 Latitude + Longitude +
                                 as.factor(State.Code) +
                                 perc_carpool_vehicle + perc_private_vehicle +
                                 total_population +
                                 ppt + s(tmax) + s(tmin) + s(tmean),
                               random = ~(1|GEOID),
                               data = train_df)

anova(gam_pol_linear_coords$mer, gam_pol_linear_ppt$mer)
```

Since our p-value is high, then the ANOVA test favors the simpler model with 60 parameters instead of the spline model with 61 parameters.

```{r nextCompare2, cache=TRUE}
gam_pol_linear_tmax <- gamm4(X50th.Percentile ~
                                 Latitude + Longitude +
                                 as.factor(State.Code) +
                                 perc_carpool_vehicle + perc_private_vehicle +
                                 total_population +
                                 ppt + tmax + s(tmin) + s(tmean),
                               random = ~(1|GEOID),
                               data = train_df)

anova(gam_pol_linear_ppt$mer, gam_pol_linear_tmax$mer)
```

Since our p-value is high, then the ANOVA test favors the simpler model with 59 parameters instead of the spline model with 60 parameters.

```{r nextCompare3, cache=TRUE}
gam_pol_linear_tmin <- gamm4(X50th.Percentile ~
                                 Latitude + Longitude +
                                 as.factor(State.Code) +
                                 perc_carpool_vehicle + perc_private_vehicle +
                                 total_population +
                                 ppt + tmax + tmin + s(tmean),
                               random = ~(1|GEOID),
                               data = train_df)

anova(gam_pol_linear_tmax$mer, gam_pol_linear_tmin$mer)
```

Since our p-value is high, then the ANOVA test favors the simpler model with 58 parameters instead of the spline model with 59 parameters.

```{r nextCompare4, cache=TRUE}
gam_pol_linear <- gamm4(X50th.Percentile ~
                                 Latitude + Longitude +
                                 as.factor(State.Code) +
                                 perc_carpool_vehicle + perc_private_vehicle +
                                 total_population +
                                 ppt + tmax + tmin + tmean,
                               random = ~(1|GEOID),
                               data = train_df)

anova(gam_pol_linear_tmin$mer, gam_pol_linear$mer)
```

Since our p-value is 1, then the ANOVA test favors the simpler model with 57 parameters instead of the spline model with 58 parameters.

### Final GAM Evaluation

```{r LMER, cache=TRUE}
# pol_linear <- lmer(X50th.Percentile ~
#                                  Latitude + Longitude +
#                                  as.factor(State.Code) +
#                                  perc_carpool_vehicle + perc_private_vehicle +
#                                  total_population +
#                                  ppt + tmax + tmin + tmean + (1|GEOID),
#                                data = train_df)
# 
# anova(gam_pol_linear_tmin$mer, pol_linear)
```


```{r GAMpreds, cache=TRUE}
# Compare "best" and all-spline models
### Train
y_hat_train_no_spline <- gam_pol_linear$gam$fitted.values
y_hat_train_splines <- gam_pol_spline_coords$gam$fitted.values

### Test
y_hat_test_no_spline <- predict(gam_pol_linear$gam,
                           newdata = (test_df %>% filter(State.Code != "28")),
                           se.fit = TRUE)
y_hat_test_splines <- predict(gam_pol_spline_coords$gam,
                         newdata = (test_df %>% filter(State.Code != "28")),
                         se.fit = TRUE)
```

```{r metricsInit, cache=TRUE}
smape <- function(y_hat, y) {
  return (mean(
      abs(y_hat - y) / ((abs(y_hat) + abs(y)) / 2),
       na.rm = TRUE) * 100)
}

rmse <- function(y_hat, y) {
  return(sqrt(mean((y - y_hat)^2, na.rm = TRUE)))
}
```

```{r}
# Get train and test metrics for both models
y_test <- (test_df %>% filter(State.Code != "28"))$X50th.Percentile

# Train
### SMAPE
cat("SMAPE (train: splines)",
    smape(y_hat_train_splines, train_df$X50th.Percentile), "\n")
cat("SMAPE (train: no splines)",
    smape(y_hat_train_no_spline, train_df$X50th.Percentile), "\n\n")

### RMSE
cat("RMSE (train: splines)",
    rmse(y_hat_train_splines, train_df$X50th.Percentile), "\n")
cat("RMSE (train: no splines)",
    rmse(y_hat_train_no_spline, train_df$X50th.Percentile), "\n\n")

# Test
### SMAPE
cat("SMAPE (test: splines)",
    smape(y_hat_test_splines$fit, y_test), "\n")
cat("SMAPE (test: no splines)",
    smape(y_hat_test_no_spline$fit, y_test), "\n\n")

### RMSE
cat("RMSE (test: splines)",
    rmse(y_hat_test_splines$fit, y_test), "\n")
cat("RMSE (test: no splines)",
    rmse(y_hat_test_no_spline$fit, y_test), "\n\n")
```

GAM with splines performs better

### GAM without covariates

```{r noCovarGAM, cache=TRUE}
gam_pol_splines_no_covar <- gamm4(X50th.Percentile ~
                                    s(Latitude) + s(Longitude),
                                  data = train_df)

y_hat_train_splines_no_covar <- gam_pol_splines_no_covar$gam$fitted.values

### Test
y_hat_test_splines_no_covar <- predict(gam_pol_splines_no_covar$gam,
                           newdata = (test_df %>% filter(State.Code != "28")),
                           se.fit = TRUE)
```

```{r evalCovarNoCovar, cache=TRUE}
# Train
### SMAPE
cat("SMAPE (train: splines covar)",
    smape(y_hat_train_splines, train_df$X50th.Percentile), "\n")
cat("SMAPE (train: splines no covar)",
    smape(y_hat_train_splines_no_covar, train_df$X50th.Percentile), "\n\n")

### RMSE
cat("RMSE (train: splines covar)",
    rmse(y_hat_train_splines, train_df$X50th.Percentile), "\n")
cat("RMSE (train: splines no covar)",
    rmse(y_hat_train_splines_no_covar, train_df$X50th.Percentile), "\n\n")

# Test
### SMAPE
cat("SMAPE (test: splines covar)",
    smape(y_hat_test_splines$fit, y_test), "\n")
cat("SMAPE (test: splines no covar)",
    smape(y_hat_test_splines_no_covar$fit, y_test), "\n\n")

### RMSE
cat("RMSE (test: splines covar)",
    rmse(y_hat_test_splines$fit, y_test), "\n")
cat("RMSE (test: splines no covar)",
    rmse(y_hat_test_splines_no_covar$fit, y_test), "\n\n")
```



