---
title: "TEMPO"
author: "Elizabeth Chun"
date: "2024-11-16"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)

library(ncdf4)
library(CFtime)
library(lattice)
library(RColorBrewer)

library(geosphere)

library(BRISC)

# plotting
library(sf)
library("rnaturalearth")
library(basemapR)
```

# Load TEMPO data

```{r}
ncin <- nc_open('final_project/C2930725014-LARC_CLOUD_merged_3nods.nc4')
print(ncin)

lat = ncvar_get(ncin,"geolocation/latitude")
lon = ncvar_get(ncin,"geolocation/longitude")
# time = ncvar_get(ncin,"geolocation/time")

no2 = ncvar_get(ncin,"product/vertical_column_troposphere")
qa_flag = ncvar_get(ncin,"product/main_data_quality_flag")

no2_df = tibble(
  lat = as.vector(lat),
  lon = as.vector(lon),
  no2 = as.vector(no2),
  qa_flag = as.vector(qa_flag)
)
```

# process NO2 data

Filter for quality checking and add distance to nearest big city

```{r}
no2_df_clean = no2_df %>%
  filter(complete.cases(.) & qa_flag == 0) %>%
  filter(no2 > 0)
```

```{r}
no2_sf = st_as_sf(no2_df_clean, coords = c("lon", "lat"), crs = 4326)
```


```{r}
# major cities
cities = tibble(
  names = c('austin', 'houston', 'san_antonio', 'el_paso', 'dallas', 'ftworth'),
  lon = c(-97.7431, -95.3701, -98.4946, -106.4850, -96.7970, -97.3308),
  lat = c(30.2672, 29.7601, 29.4252, 31.7619, 32.7767, 32.7555)
) 

cities_sf = st_as_sf(cities, coords = c("lon", "lat"), crs = 4326)
```


```{r}
redo = FALSE
if (redo) {
  no2_dist = no2_df_clean

  for (city in cities$names) {
    citycenter = cities %>%
      filter(names == city)
    no2_dist[city] = distGeo(citycenter[, 2:3], no2_dist[, c("lon", "lat")])
  }
  
  no2_dist_city = no2_dist %>%
    rowwise() %>%
    mutate(
      dist_to_city = min(across(austin:ftworth)),
      city = cities$names[which.min(across(austin:ftworth))]
    ) %>%
    ungroup()
} else {
  no2_dist_city = readRDS("final_project/no2_dist_city.rds")
}

```

```{r}
no2_dist_city = no2_dist_city %>%
  mutate(
    # multiply by rough height of column to get molecules/cubic cm
    # divide by average number of molecules/cubic cm 
    ## both conversions very rough
    no2_ppm = no2*12000 / (2.46*10^19)
  ) 

tx_xform = sf::st_transform(tx, crs = 4326)
no2_dist_city_sf = st_as_sf(no2_dist_city, coords = c("lon", "lat"), crs = 4326)
no2_df_final_sf = no2_dist_city_sf[tx_xform, ]

no2_df_final = no2_df_final_sf %>% 
  st_drop_geometry() %>%  
  bind_cols(st_coordinates(no2_df_final_sf)) %>%
  select(lon = X, lat = Y, no2_ppm, dist_to_city, city, austin, houston, san_antonio, el_paso, dallas, ftworth)
```



```{r}
# try to read highways
highways = sf::st_read('final_project/highways/Texas_National_Highway_Freight_Network.shp')
```


```{r}
bbox_no2 = st_bbox(no2_df_final_sf)

ggplot() + 
  base_map(bbox_no2, increase_zoom = 3, basemap = 'mapnik') +
  geom_sf(data = no2_df_final_sf[sample(nrow(no2_df_final_sf), 20000), ],
          aes(color = no2_ppm), size = 2) +
  scale_color_gradient2(low = '#00FF0010', mid = '#FFA500CC', high = "red",
                       limits = c(0, 10), na.value = NA, midpoint = 4) +
  geom_sf(data = highways, linewidth = 0.25, color = 'grey40')
  # scale_color_gradientn(colours = c("green", 'green', 'green', 'orange', 'red'))
```

# data split - random

```{r data splitting}
### data splitting
set.seed(27)

# get 80/20 train test split
train_idx = sample(nrow(no2_df_final), 0.8*nrow(no2_df_final))
train_all = no2_df_final[train_idx, ]
test_all = no2_df_final[-train_idx, ]
```

```{r}
st_as_sf(test_all, coords = c("lon", "lat"), crs = 4326) %>%
  ggplot() + 
    base_map(bbox_no2, increase_zoom = 4, basemap = 'mapnik') +
    geom_sf(aes(color = no2_ppm), alpha = 0.75) +
    scale_color_gradient2(low = 'white', mid = '#FF474D', high = "red", 
                         limits = c(0, 35), na.value = NA, midpoint = 18)
```



# 3.2 Nearest neighbors GP

Now I will fit an NNGP on the entire training set with neighbors using the BRISC package.

```{r fit nngp}
fit_nngp = BRISC_estimation(coords = as.matrix(train_all[, c('lon', 'lat')]), 
                            x = cbind(1, as.matrix(
                              train_all[, c('dist_to_city', 'lon', 'lat')])), 
                            y = train_all$no2_ppm, 
                            cov.model = 'exponential', n.neighbors = 20)
print(fit_nngp$Beta)
print(fit_nngp$Theta)
```



# 4. Model results and predictions

## 4.2 Nearest neighbors GP

For the NNGP, here are the estimated parameters. Note that BRISC reports $\phi$ as the reciprocal of the standard definition so I have taken $\frac{1}{\phi}$.

$\hat{\beta}$'s = [`r fit_nngp$Beta`]  
$\hat{\sigma}^2$ = `r fit_nngp$Theta[1]`  
$\phi$ = `r 1/fit_nngp$Theta[3]`  
$\hat{\tau}^2$ = `r fit_nngp$Theta[2]`    


```{r nngp prediction}
preds_nngp = BRISC_prediction(fit_nngp, coords.0 = as.matrix(test_all[, c('lon', 'lat')]), 
                              X.0 = cbind(1, as.matrix(
                                test_all[, c('dist_to_city', 'lon', 'lat')])))
rmse_nngp = sqrt(mean((test_all$no2_ppm - preds_nngp$prediction)^2))
mape = mean(abs(test_all$no2_ppm - preds_nngp$prediction)/test_all$no2_ppm*100)

resids = test_all$no2_ppm - preds_nngp$prediction

print(rmse_nngp)
print(mape)

hist(resids)
hist(preds_nngp$prediction)
```

```{r}
# plot random predictions map

pred_df_rand = test_all[, c('lon', 'lat')] %>%
  mutate(
    no2_ppm_pred = preds_nngp$prediction
  ) 
pred_df_rand_sf = st_as_sf(pred_df_rand, coords = c("lon", "lat"), crs = 4326)

ggplot(pred_df_rand_sf) + 
  base_map(bbox_no2, increase_zoom = 4, basemap = 'mapnik') +
  geom_sf(aes(color = no2_ppm_pred), alpha = 0.75) +
  scale_color_gradient2(low = 'white', mid = '#FF474D', high = "red", 
                       limits = c(0, 34), na.value = NA, midpoint = 17)
```


# data splitting

```{r data splitting}

# get holdout city test split
train_idx = which(no2_df_final$city != 'ftworth')
train_all = no2_df_final[train_idx, ]
test_all = no2_df_final[-train_idx, ]
```

```{r}
st_as_sf(test_all, coords = c("lon", "lat"), crs = 4326) %>%
  ggplot() + 
    base_map(bbox_no2, increase_zoom = 4, basemap = 'mapnik') +
    geom_sf(aes(color = no2_ppm), alpha = 0.75) +
    scale_color_gradient2(low = 'white', mid = '#FF474D', high = "red", 
                         limits = c(0, 35), na.value = NA, midpoint = 18)
```


## 3.2 Nearest neighbors GP

Now I will fit an NNGP on the entire training set with neighbors using the BRISC package.

```{r fit nngp}
fit_nngp2 = BRISC_estimation(coords = as.matrix(train_all[, c('lon', 'lat')]), 
                            x = cbind(1, as.matrix(train_all[, c('lon', 'lat')])), 
                            y = train_all$no2_ppm, 
                            cov.model = 'exponential', n.neighbors = 20)
print(fit_nngp2$Beta)
print(fit_nngp2$Theta)
```



# 4. Model results and predictions

## 4.2 Nearest neighbors GP

For the NNGP, here are the estimated parameters. Note that BRISC reports $\phi$ as the reciprocal of the standard definition so I have taken $\frac{1}{\phi}$.

$\hat{\beta}$'s = [`r fit_nngp$Beta`]  
$\hat{\sigma}^2$ = `r fit_nngp$Theta[1]`  
$\phi$ = `r 1/fit_nngp$Theta[3]`  
$\hat{\tau}^2$ = `r fit_nngp$Theta[2]`    


```{r nngp prediction}
preds_nngp2 = BRISC_prediction(fit_nngp2, coords.0 = as.matrix(test_all[,  c('lon', 'lat')]), 
                              X.0 = cbind(1, as.matrix(test_all[,  c('lon', 'lat')])))
rmse_nngp2 = sqrt(mean((test_all$no2_ppm - preds_nngp2$prediction)^2))
mape2 = mean(abs(test_all$no2_ppm - preds_nngp2$prediction)/test_all$no2_ppm*100)

resids2 = test_all$no2_ppm - preds_nngp2$prediction

print(rmse_nngp2)
print(mape2)
hist(resids2)
hist(preds_nngp2$prediction)
```


```{r}
# plot random predictions map

pred_df_rand = test_all[, c('lon', 'lat')] %>%
  mutate(
    no2_ppm_pred = preds_nngp2$prediction
  ) 
pred_df_rand_sf = st_as_sf(pred_df_rand, coords = c("lon", "lat"), crs = 4326)

ggplot(pred_df_rand_sf) + 
  base_map(bbox_no2, increase_zoom = 4, basemap = 'mapnik') +
  geom_sf(aes(color = no2_ppm_pred), alpha = 0.75) +
  scale_color_gradient2(low = 'white', mid = '#FF474D', high = "red", 
                       limits = c(0, 34), na.value = NA, midpoint = 17)
```

